---
title: "BRFSS Healthcare Access Analysis"
author: "Brandon Rugg"
date: "2026-04-02"
output: html_notebook
---
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(dplyr)
library(survey)

source("R/utils.R")
```

Dataset and documentation available at https://www.cdc.gov/brfss/annual_data/annual_2024.html

```{r read_data}
# Import the 2024 BRFSS dataset (SAS Transport Format)
brfss_data <- read_xpt("data/LLCP2024.XPT")

# Generate summary table of all variables. Labels are truncated-- the full titles are available in the 2024 BRFSS Codebook CDC.
var_dict <- build_variable_dictionary(brfss_data)
head(var_dict)
```

```{r}
# List of columns to keep
keep_vars <- c("_STATE", "_LLCPWT", "_PSU", "_STSTR", 
               "_AGEG5YR", "_SEX", "_RACEGR3", "_URBSTAT",
               "_FLSHOT7", "_PNEUMO3", "_AIDTST4", "_CRVSCRN",
               "_BMI5CAT", "_RFSMOK3")

# Create smaller dataframe for more efficient analysis
brfss_small <- brfss_data[, keep_vars]
var_dict_small <- build_variable_dictionary(brfss_small)
```
---
## Analysis of flu shot coverage for those > 65 years of age

Variable `_FLSHOT7` already only includes those 65+. A response of '9' indicates 'Don't know/Not Sure' Or 'Refused/Missing'. Dropped for analysis.
```{r}
# Clean the flu shot variable and remove missing codes
flu_data <- brfss_small[, c("_FLSHOT7", "_PSU", "_STSTR", "_LLCPWT", "_STATE")]
flu_data$flu_shot_clean <- ifelse(flu_data$`_FLSHOT7` %in% c(9, NA), NA, flu_data$`_FLSHOT7`)

# Calculate percentage of '9's in each state to make sure state-by-state comparison is valid after dropping
pct_9_by_state <- brfss_small %>%
  group_by(`_STATE`) %>%
  summarize(
    n_total = n(),
    n_9 = sum(`_FLSHOT7` == 9, na.rm = TRUE),
    pct_9 = n_9 / n_total * 100
  ) %>%
  arrange(desc(pct_9))  # sort from highest to lowest

pct_9_by_state
```

% of respondents answering "9" for `_FLSHOT7` varies between 8.05 and 1.25% state-by-state. As it's relatively low, it seems unlikely that it would meaningfully bias interstate comparisons.

```{r}
# Adjust for "lonely" PSUs (strata with only one PSU) in variance estimation
options(survey.lonely.psu = "adjust")

# Define the survey design for flu vaccination data
flu_design <- svydesign(
  id = ~`_PSU`,           # Primary Sampling Unit
  strata = ~`_STSTR`,     # Stratification variable
  weights = ~`_LLCPWT`,   # Survey weights
  data = flu_data,
  nest = TRUE            # Indicates nested structure
)

# Estimate the mean flu vaccination rate
flu_est <- svymean(~factor(flu_shot_clean), design = flu_design, na.rm = TRUE)

flu_est
```
```{r}
# Format results for readability
flu_yes <- coef(flu_est)[1]
flu_ci  <- confint(flu_est)[1, ]

cat(sprintf(
  "Flu vaccination rate (age 65+): %.1f%% (95%% CI: %.1f%%–%.1f%%)\n",
  flu_yes * 100,
  flu_ci[1] * 100,
  flu_ci[2] * 100
))
```

