---
title: "BRFSS Healthcare Access Analysis"
author: "Brandon Rugg"
date: "2026-04-02"
output: html_notebook
---
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(survey)
library(knitr)
library(kableExtra)
library(usmap)
library(ggplot2)
library(scales)

source("R/utils.R")

if (!dir.exists("outputs")) dir.create("outputs")
```

Dataset and documentation available at https://www.cdc.gov/brfss/annual_data/annual_2024.html

```{r read_data}
# Import the 2024 BRFSS dataset (SAS Transport Format)
brfss_data <- read_xpt("data/LLCP2024.XPT")

# Generate summary table of all variables. Labels are truncated-- the full titles are available in the 2024 BRFSS Codebook CDC.
var_dict <- build_variable_dictionary(brfss_data)
head(var_dict)
```

```{r}
# List of columns to keep
keep_vars <- c("_STATE", "_LLCPWT", "_PSU", "_STSTR", 
               "_AGEG5YR", "_SEX", "_RACEGR3", "_URBSTAT",
               "_FLSHOT7", "_PNEUMO3", "_AIDTST4", "_CRVSCRN",
               "_BMI5CAT", "_RFSMOK3")

# Create smaller dataframe for more efficient analysis
brfss_small <- brfss_data[, keep_vars]
var_dict_small <- build_variable_dictionary(brfss_small)
```
```{r}
# Optional after creating the smaller working dataset
rm(brfss_data)      # Remove the large original
gc()                # Run garbage collection to free memory
```

---
## Analysis of flu shot coverage for those > 65 years of age

Variable `_FLSHOT7` already only includes those 65+. A response of '9' indicates 'Don't know/Not Sure' Or 'Refused/Missing'. Dropped for analysis.
```{r}
# Clean the flu shot variable and remove missing codes
brfss_small$flu_shot_clean <- ifelse(brfss_small$`_FLSHOT7` %in% c(9, NA), NA, brfss_small$`_FLSHOT7`)

# Calculate percentage of '9's in each state to make sure state-by-state comparison is valid after dropping
pct_9_by_state <- brfss_small %>%
  group_by(`_STATE`) %>%
  summarize(
    n_total = n(),
    n_9 = sum(`_FLSHOT7` == 9, na.rm = TRUE),
    pct_9 = n_9 / n_total * 100
  ) %>%
  arrange(desc(pct_9))  # sort from highest to lowest

pct_9_by_state
```

% of respondents answering "9" for `_FLSHOT7` varies between 8.05 and 1.25% state-by-state. As it's relatively low, it seems unlikely that it would meaningfully bias interstate comparisons.

```{r}
# Adjust for "lonely" PSUs (strata with only one PSU) in variance estimation
options(survey.lonely.psu = "adjust")

# Define the survey design for flu vaccination data
flu_design <- svydesign(
  id = ~`_PSU`,           # Primary Sampling Unit
  strata = ~`_STSTR`,     # Stratification variable
  weights = ~`_LLCPWT`,   # Survey weights
  data = brfss_small,
  nest = TRUE            # Indicates nested structure
)

# Estimate the mean flu vaccination rate
flu_est <- svymean(~factor(flu_shot_clean), design = flu_design, na.rm = TRUE)

flu_est
```
```{r}
# Format results for readability
flu_yes <- coef(flu_est)[1]
flu_ci  <- confint(flu_est)[1, ]

cat(sprintf(
  "Flu vaccination rate (age 65+): %.1f%% (95%% CI: %.1f%%–%.1f%%)\n",
  flu_yes * 100,
  flu_ci[1] * 100,
  flu_ci[2] * 100
))
```
```{r}
flu_by_state <- svyby(
  ~factor(flu_shot_clean),
  ~`_STATE`,
  design = flu_design,
  FUN = svymean,
  na.rm = TRUE
)
```


```{r}
# Map state codes to names
state_labels <- c(
  `1` = "Alabama", `2` = "Alaska", `4` = "Arizona", `5` = "Arkansas",
  `6` = "California", `8` = "Colorado", `9` = "Connecticut", `10` = "Delaware",
  `11` = "District of Columbia", `12` = "Florida", `13` = "Georgia", `15` = "Hawaii",
  `16` = "Idaho", `17` = "Illinois", `18` = "Indiana", `19` = "Iowa",
  `20` = "Kansas", `21` = "Kentucky", `22` = "Louisiana", `23` = "Maine",
  `24` = "Maryland", `25` = "Massachusetts", `26` = "Michigan", `27` = "Minnesota",
  `28` = "Mississippi", `29` = "Missouri", `30` = "Montana", `31` = "Nebraska",
  `32` = "Nevada", `33` = "New Hampshire", `34` = "New Jersey", `35` = "New Mexico",
  `36` = "New York", `37` = "North Carolina", `38` = "North Dakota", `39` = "Ohio",
  `40` = "Oklahoma", `41` = "Oregon", `42` = "Pennsylvania", `44` = "Rhode Island",
  `45` = "South Carolina", `46` = "South Dakota", `48` = "Texas", `49` = "Utah",
  `50` = "Vermont", `51` = "Virginia", `53` = "Washington", `54` = "West Virginia",
  `55` = "Wisconsin", `56` = "Wyoming", `66` = "Guam", `72` = "Puerto Rico",
  `78` = "Virgin Islands"
)

# Convert state variable in flu_by_state to the labels
flu_by_state$State <- state_labels[as.character(flu_by_state$`_STATE`)]

flu_by_state
```

```{r}
# 95% CI using normal approximation
flu_by_state <- flu_by_state %>%
  mutate(
    flu_rate = `factor(flu_shot_clean)1`,              # your column with vaccination rate (1 = yes)
    flu_se = `se.factor(flu_shot_clean)1`,                    # your column with SE
    lower_ci = flu_rate - 1.96 * flu_se,
    upper_ci = flu_rate + 1.96 * flu_se
  )

# Create table of state flu vaccination data
flu_by_state %>%
  select(State, flu_rate, lower_ci, upper_ci) %>%
  arrange(desc(flu_rate)) %>%
  kable(digits = 3, 
        col.names = c("State", "Vaccination Rate", "Lower CI", "Upper CI"),
        caption = "Flu Vaccination Rates (Age 65+) by State with 95% CIs") %>%
  kable_styling(full_width = FALSE)
```

```{r}
flu_by_state <- flu_by_state %>%
  mutate(state = state.abb[match(State, state.name)])

flu_by_state <- flu_by_state %>%
  mutate(flu_rate = `factor(flu_shot_clean)1`)

plot_usmap(
  data = flu_by_state,
  values = "flu_rate",
  regions = "states"
) +
  scale_fill_viridis_c(
    labels = percent_format(accuracy = 1),
    name = "Flu Vaccination Rate"
  ) +
  theme(legend.position = "right") +
  labs(title = "Flu Vaccination Rate (Age 65+) by State")

ggsave("outputs/flu_map.png", width = 10, height = 6, dpi = 300)
```
*Note: Tennessee is absent from the 2024 BRFSS public dataset. [CDC BRFSS 2024 documentation](https://www.cdc.gov/brfss/annual_data/annual_2024.html).*

### Comparison of flu vaccination rates (65+) for urban (1) vs rural (2) populations
```{r}
# Estimate mean flu vaccination by urban/rural
flu_urban_est <- svyby(
  ~factor(flu_shot_clean),
  ~`_URBSTAT`,
  design = flu_design,
  FUN = svymean,
  na.rm = TRUE
)

# Format results
flu_urban_est <- flu_urban_est %>%
  mutate(
    flu_rate = `factor(flu_shot_clean)1`,      # vaccination = 1 (yes)
    flu_se = `se.factor(flu_shot_clean)1`,
    lower_ci = flu_rate - 1.96 * flu_se,
    upper_ci = flu_rate + 1.96 * flu_se,
    urban_rural = ifelse(`_URBSTAT` == 1, "Urban", "Rural")
  )

# Display
flu_urban_est[, c("urban_rural", "flu_rate", "lower_ci", "upper_ci")]
```
### Comparison of flu vaccination rates (65+) by ethnicity
```{r}
# Weighted mean by race (65+), ignoring 9 (Don't know/Refused)
flu_race_est <- svyby(
  ~factor(flu_shot_clean),
  ~factor(ifelse(`_RACEGR3` %in% c(1,2,3,4,5), `_RACEGR3`, NA)),
  design = flu_design,
  FUN = svymean,
  na.rm = TRUE
)

# Official 5-level race/ethnicity labels
race_labels <- c(
  "White only, Non-Hispanic",
  "Black only, Non-Hispanic",
  "Other race only, Non-Hispanic",
  "Multiracial, Non-Hispanic",
  "Hispanic"
)

# Explicit column names for "Yes" and its SE
rate_col <- "factor(flu_shot_clean)1"
se_col   <- "se.factor(flu_shot_clean)1"

# Print percentages with 95% CIs
for (i in seq_along(race_labels)) {
  rate <- flu_race_est[[rate_col]][i]
  se   <- flu_race_est[[se_col]][i]
  lower <- rate - 1.96 * se
  upper <- rate + 1.96 * se
  
  cat(sprintf(
    "Flu vaccination rate (%s, 65+): %.1f%% (95%% CI: %.1f%%–%.1f%%)\n",
    race_labels[i],
    rate * 100,
    lower * 100,
    upper * 100
  ))
}
```
```{r}
# Inspect column names once
names(flu_race_est)
```
### Comparison of flu vaccination rates (65+) by sex
```{r}
# Adjust for "lonely" PSUs (strata with only one PSU) in variance estimation
options(survey.lonely.psu = "adjust")

# Estimate mean flu vaccination by male (1) or female (2)
flu_sex_est <- svyby(
  ~factor(flu_shot_clean),
  ~`_SEX`,
  design = flu_design,
  FUN = svymean,
  na.rm = TRUE
)

# Format results
flu_sex_est <- flu_sex_est %>%
  mutate(
    flu_rate = `factor(flu_shot_clean)1`,      # vaccination = 1 (yes)
    flu_se = `se.factor(flu_shot_clean)1`,
    lower_ci = flu_rate - 1.96 * flu_se,
    upper_ci = flu_rate + 1.96 * flu_se,
    male_female = case_when(
      `_SEX` == 1 ~ "Male",
      `_SEX` == 2 ~ "Female",
      TRUE ~ NA_character_
    )
  )

# Display
flu_sex_est[, c("male_female", "flu_rate", "lower_ci", "upper_ci")]
```



